---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute 
intervals through out the day. The data consists of two months of data from an anonymous individual collected during 
the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data
The data for this assignment can be downloaded from the course web site:
Dataset: Activity monitoring data [52K] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this dataset are:
steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

```{r setup, include=FALSE, cache=FALSE}
options(scipen = 1, digits = 2) #set to two decimal 
```

Load library
```{r Load Library}
library(ggplot2)
```

Define some global variables
```{r Define global variables}
cURLActivity <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
cZipActivity <- "activity.zip"
cFileActivity <- "activity.csv"
```

Download the data file from the given URL
```{r Download file}
if (!file.exists(cZipActivity)){
  download.file(cURLActivity, cZipActivity, method="curl")
}  
```

Unzip the data file if the zip file exists but the uncompressed file does not exist
```{r Unzip file}
if (file.exists(cZipActivity) & !file.exists(cFileActivity)) { 
  unzip(cZipActivity)
}
```

Read the activity data into dataframe
```{r Read data file into dataframe}
if (!file.exists(cFileActivity)) {
  print("Required data file is not ready. Please check the zip data file.")
  return
} else {
  dfActivity <- read.csv(cFileActivity, colClasses=c("integer","Date","character"))
}
```


## What is mean total number of steps taken per day?
* Ignore the missing values in the dataset.
* Make a histogram of the total number of steps taken each day
```{r}
dfTotalStepsEachDay <- aggregate(steps ~ date, dfActivity, sum)
g <- ggplot(dfTotalStepsEachDay, aes(steps))
p <- g + geom_histogram(bins=5, color="black", fill="grey")
p <- p + ggtitle("Histogram of the total number of steps taken each day")+xlab("Total number of steps each day")
p <- p + theme(plot.title = element_text(hjust = 0.5))
print(p)
```

* Calculate and report the mean and median total number of steps taken per day
```{r}
nMeanStepsEachDay <- mean(dfTotalStepsEachDay$steps)
nMedianStepsEachDay <- median(dfTotalStepsEachDay$steps)
```
The mean of total number of steps taken per day is `r nMeanStepsEachDay`.

The median of total number of steps taken per day is `r nMedianStepsEachDay`.

The mean and the median are so close as shown in the histogram below:
```{r}
p <- p + geom_vline(aes(xintercept=nMeanStepsEachDay, color="Mean"), linewidth=1.5)
p <- p + geom_vline(aes(xintercept=nMedianStepsEachDay, color="Median"), linetype="dashed", linewidth=1.5)
p <- p + scale_color_manual(name = "Statistics", values = c(Mean="red", Median="blue"))
print(p)
```

## What is the average daily activity pattern?
* Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days 
```{r}
dfStepsInterval <- aggregate(steps~interval, dfActivity, FUN=mean)
dfStepsInterval <- transform(dfStepsInterval, interval=as.numeric(interval))
g <- ggplot(dfStepsInterval, aes(interval, steps))
p <- g+geom_line()
p <- p + ggtitle("Average Daily Activity Pattern")+xlab("5-minute Interval")+ylab("Average Number of Steps Taken")
p <- p + theme(plot.title = element_text(hjust = 0.5))
print(p)
```

* Find out which 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps
```{r}
nIntervalWithMaxNoOfSteps <- dfStepsInterval$interval[which.max(dfStepsInterval$steps)]
```
The 5-minute interval that contains the maximum number of steps is `r nIntervalWithMaxNoOfSteps`.


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
